---
title: "Igor Kuivjogi Fernandes"
subtitle: "Assignment 3"
# date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Part A (Answers)

## i) 

$n = 33$ and $p = 1200$.   

## ii)

| Simulated data: summary| Actual_data|
|-----------------------:|-----------:|
|                  50.754|     541.154|
|                  48.923|     146.995|
|                  47.647|      92.439|
|                  46.576|      83.786|
|                  45.637|      46.546|
|                  44.827|      35.040|
|                  44.006|      30.305|
|                  43.222|      26.793|
|                  42.506|      20.770|
|                  41.819|      18.052|
|                  41.164|      16.227|
|                  40.487|      11.853|
|                  39.900|      10.903|
|                  39.294|      10.630|
|                  38.645|       9.967|
|                  38.039|       8.989|
|                  37.462|       8.864|
|                  36.880|       8.419|
|                  36.281|       7.909|
|                  35.711|       7.236|
|                  35.121|       6.831|
|                  34.537|       6.318|
|                  33.950|       5.980|
|                  33.415|       5.548|
|                  32.789|       5.070|
|                  32.204|       4.956|
|                  31.621|       4.527|
|                  31.011|       4.318|
|                  30.346|       3.970|
|                  29.665|       3.485|
|                  28.939|       3.368|
|                  28.062|       2.753|
|                   0.000|       0.000|

As summary of simulated eigenvalues becomes larger than the eigenvalues from the data at the $6^{th}$ row, we should choose $k = 5$ factors.   

## iii)

  a)   
  
  |               |        |
  |:--------------|-------:|
  |var_L2_initial |   0.062|
  |var_L2         |   0.180|
  |ratio          |   2.907|
  
  b)    
  
  |                      |     1|     2|     3|     4|     5|
  |:---------------------|-----:|-----:|-----:|-----:|-----:|
  |var_L2_initial_factor | 0.028| 0.017| 0.009| 0.006| 0.003|
  |var_L2_factor         | 0.034| 0.058| 0.017| 0.058| 0.013|
  |ratio_factor          | 1.241| 3.345| 1.987| 9.962| 4.873|
  
  c)    
  
  The fourth factor showed the maximum improvement due to varimax, whereas the first factor showed the minimum improvement.
  
  
## iv)

How many variables have at most 30% variability explained by factors: $5$ variables.

The variable which is worst explained by the factors: $684^{th}$ variable.

Communality of $236^{th}$ variable: $\approx 0.6933$.

Proportion of overall variability explained by $2^{nd}$ factor: $\approx 0.1844$.

Correlation between $23^{rd}$ and $317^{th}$ variables: $\approx 0.50$.

Specific variance of $840^{th}$ variable: $\approx 0.2964$.


## v)

$$
\hat{f}^{(16)} = \begin{bmatrix}
  -0.42722879 \\
  -0.09308803 \\
  1.77837205 \\
  -0.58302362 \\
  0.49855498 \\
\end{bmatrix}
$$

# Part B (code)

```{r}
# i)
tab <- as.matrix(read.table("Factor_application_data.txt", header = F))
n <- nrow(tab)
p <- ncol(tab)
```


```{r, results = 'hide'}
# ii)
for (i in 1:p) tab[ , i] <- (tab[,i] - mean(tab[, i])) / sd(tab[, i])

B <- matrix(0, p, n)
for (i in 1:n) B[, i] <- tab[i, ] - 0  # data is already standardized

if (n < p) {
  B_transpose_B <- t(B) %*% B
  eigenvalues_data <- eigen(B_transpose_B, only.values = T)$values / (n - 1)
} else {
  B_B_transpose <- B %*% t(B)
  eigenvalues_data <- eigen(B_B_transpose, only.values = T)$values / (n - 1)
}

# parallel analysis
sim_no <- 2000
eigenvalue_store <- matrix(0, min(n, p), sim_no)
set.seed(123)
for (j in 1:sim_no) {
  data_sim <- matrix(rnorm(n * p), n, p)
  for (i in 1:p) data_sim[ ,i] <- (data_sim[,i] - mean(data_sim[, i])) / sd(data_sim[, i])
  B <- matrix(0, p, n)
  for (i in 1:n) {
    B[,i] <- data_sim[i, ] - 0  # already standardized
  }
  if (n < p) {
    B_transpose_B <- t(B) %*% B
    eigenvalues_sim <- eigen(B_transpose_B, only.values = T)$values / (n - 1)
  } else {
    B_B_transpose <- B %*%t (B)
    eigenvalues_sim <- eigen(B_B_transpose, only.values = T)$values / (n - 1)
  }
  eigenvalue_store[, j] <- eigenvalues_sim
}

quantile_choice <- 0.95
summary_simulation <- apply(eigenvalue_store, 1, quantile, p = quantile_choice)
## if mean is used
# summary_simulation <- apply(eigenvalue_store, 1, mean)
show_comparison <- cbind(summary_simulation, eigenvalues_data)
colnames(show_comparison) <- c("Simulated data: summary", "Actual_data")
# knitr::kable(round(show_comparison, 3), format = "markdown")
k <- which(show_comparison[, 1] > show_comparison[, 2])[1] - 1
```


```{r, results = 'hide'}
# iii)
B <- matrix(0, p, n)
for (i in 1:n) B[, i] <- tab[i, ] - 0 # data is already standardized
if (n < p) {
  B_transpose_B <- t(B) %*% B
  eigen_results <- eigen(B_transpose_B)
  eigen_kvalues <- eigen_results$values[1:k]
  eigen_kvectors <- eigen_results$vectors[, 1:k]
  eigvec_Sigma_hat_k <- matrix(0, p, k)
  for (i in 1:k) {
    eigvec_Sigma_hat_k[ ,i] = c(1 / sqrt(eigen_kvalues[i])) * 
      (B %*% matrix(eigen_kvectors[, i], ncol = 1))
  }
  eigval_Sigma_hat_k <- eigen_kvalues / (n - 1)
} else {
  B_B_transpose <- B% *% t(B)
  eigen_results <- eigen(B_B_transpose)
  eigvec_Sigma_hat_k <- eigen_results$vectors[, 1:k]
  eigval_Sigma_hat_k <- eigen_results$values[1:k] / (n - 1)
}

L_initial <- eigvec_Sigma_hat_k %*% diag(sqrt(eigval_Sigma_hat_k))
L <- varimax(L_initial, normalize = F)$loadings
# print(L, cutoff = 0.01)

# a)
var_L2_initial <- sum(apply(L_initial ^ 2, 2, var))
var_L2 <- sum(apply(L ^ 2, 2, var))
ratio <- var_L2 / var_L2_initial
var_comparison <- rbind(var_L2_initial, var_L2, ratio)
# knitr::kable(round(var_comparison, 3), format = "markdown")  # almost 3x

# b)
var_L2_initial_factor <- apply(L_initial ^ 2, 2, var)
var_L2_factor <- apply(L ^ 2, 2, var)
ratio_factor <- var_L2_factor / var_L2_initial_factor
var_comparison_factor <- rbind(var_L2_initial_factor, var_L2_factor, ratio_factor)
colnames(var_comparison_factor) <- 1:5
# knitr::kable(round(var_comparison_factor, 3), format = "markdown")

# c)
which.max(ratio_factor)
which.min(ratio_factor)
```


```{r, results = 'hide'}
# iv) 
sum_squares_row <- apply(L ^ 2, 1, sum)
length(which(sum_squares_row <= 0.3))  # 5
which.min(sum_squares_row)  # 684
sum_squares_row[236]  # 0.6933

sum_squares_col <- apply(L ^ 2, 2, sum)
sum_squares_col[2] / p  # 0.184456
L[23, ] %*% L[317, ]  # 0.50
1 - sum_squares_row[840]  # 0.2964
```


```{r, results = 'hide'}
# v)
D <- diag(1 - sum_squares_row)
Dinv <- diag(1 / diag(D))
LtDinv <- t(L) %*% Dinv
fhat_16 <- solve(LtDinv %*% L) %*% LtDinv %*% matrix(tab[16, ], ncol = 1)
```

